Draft_Application_Stochastic_Volatility

### 2021-08-10 Real Dataset SV.Rmd
Aggiunta grafici

We now offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the time-varying volatility of the observations, obtained through the application of the different filtering techniques explored in the theoretical analysis. Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows. For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$. Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.} \

First, a general point is that the sequence of percentage log returns are different across the three indices. The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents (SEE SECTION X). In particular, as previously noticed, there are extensive similarities between the DOW JONES Index and the S&P500, which can be visualized when looking at the percentage of log returns in the corresponding plots, since both are describing the US Economy. However, the STOXX50E differs quite extensively, especially in the different sensitivity to the Covid-19 Crisis and its consequences,  representing the European Economy. Aside from this note, the considerations expressed below on the sequences "filtered volatilities"\footnote{Per ora nei grafici c'è scritto "Filtered States", verrà corretto.} for the various techniques hold almost uniformly for all the three indices. Therefore, in the next paragraph we will be referring to the set of plots pertaining to one specific index (which could be any of the three).\

@@

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Molto probabilmente la frase formulata così non è formalmente corretta...bisognerebbe fare riferimento alla sezione sulla convergenza. Inoltre, va chiarito il punto della "doppia convergenza".} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500. Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}

We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark. In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles. Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below. Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm. By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never). In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}

@@

As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse. With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.\
As it turns out, such suggestion is confirmed by running the algorithms with N=1000. Passing from 100 to 1000 particles more than halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a shrinking hedge over the Auxiliary PF, which shows most impressive gains when passing from 100 to 1000 particles. Further sizable gains for every filter, except for the SIS algorithm, are observed when increasing the number of particles to 5000, a level at which both the RMSE and the MAE agree on a slight hedge of the Guided and the Bootstrap filters over the Auxiliary and the Basic PFs.\
Nevertheless, we would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others, a figure which visual inspection of the graphs below can confirm. Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (note that the Basic PF is preferred to the APF only once, by the RMSE), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles. Moreover, the RMSE and the MAE show infrequent disagreement, delivering a rather clear picture of the comparison.

>>>>

We now offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the time-varying volatility of the observations, obtained through the application of the different filtering techniques explored in the theoretical analysis. Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows. For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t/2}$. Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.} \

```{r echo=FALSE, fig.align='center', fig.cap="Observations and filtered volatility, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}

# Filtered Volatility Plots

realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
y1<-realdatasetSPX[,2]
timeframe<-c(1:length(y1))

sv_bpf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,1]
sv_lw1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 4))[,1]
sv_bapf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,1]
dfsv1<-data.frame(timeframe,y1,sv_bpf1)
dfsv2<-data.frame(timeframe,y1,sv_gpfopt1)
dfsv3<-data.frame(timeframe,y1,sv_apf1)
dfsv4<-data.frame(timeframe,y1,sv_lw1)
dfsv5<-data.frame(timeframe,y1,sv_bapf1)
dfsv6<-data.frame(timeframe,y1,sv_sis1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot2<-function(dataframe,title){
  mean=sv_gpfopt1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot3<-function(dataframe,title){
  mean=sv_apf1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot4<-function(dataframe,title){
  mean=sv_lw1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot5<-function(dataframe,title){
  mean=sv_bapf1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot6<-function(dataframe,title){
  mean=sv_sis1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}

plotobs1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
plotobs2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
plotobs3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
plotobs4<-Filtplot4(dfsv4,"Liu and West Filter SP500")
plotobs5<-Filtplot5(dfsv5,"Always Resampling SP500")
plotobs6<-Filtplot6(dfsv6,"No Resampling SP500")

ggarrange(plotobs1,plotobs2)
ggarrange(plotobs3,plotobs4)
ggarrange(plotobs5,plotobs6)

realdatasetSTOXX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
y2<-realdatasetSTOXX[,2]
timeframe<-c(1:length(y2))

sv_bpf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 3))[,1]
sv_lw2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 4))[,1]
sv_bapf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 6))[,1]
dfsv7<-data.frame(timeframe,y2,sv_bpf2)
dfsv8<-data.frame(timeframe,y2,sv_gpfopt2)
dfsv9<-data.frame(timeframe,y2,sv_apf2)
dfsv10<-data.frame(timeframe,y2,sv_lw2)
dfsv11<-data.frame(timeframe,y2,sv_bapf2)
dfsv12<-data.frame(timeframe,y2,sv_sis2)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot8<-function(dataframe,title){
  mean=sv_gpfopt2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot9<-function(dataframe,title){
  mean=sv_apf2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot10<-function(dataframe,title){
  mean=sv_lw2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot11<-function(dataframe,title){
  mean=sv_bapf2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot12<-function(dataframe,title){
  mean=sv_sis2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}

plotobs7<-Filtplot7(dfsv7,"Bootstrap Particle Filter STOXX50E")
plotobs8<-Filtplot8(dfsv8,"Opt Ker Guided Particle Filter STOXX50E")
plotobs9<-Filtplot9(dfsv9,"Auxiliary Particle Filter STOXX50E")
plotobs10<-Filtplot10(dfsv10,"Liu and West Filter STOXX50E")
plotobs11<-Filtplot11(dfsv11,"Always Resampling STOXX50E")
plotobs12<-Filtplot12(dfsv12,"No Resampling STOXX50E")

ggarrange(plotobs7,plotobs8)
ggarrange(plotobs9,plotobs10)
ggarrange(plotobs11,plotobs12)

realdatasetDJI<-read.xlsx("Dataset 17-21.xlsx", sheet = 3)
y3<-realdatasetDJI[,2]
timeframe<-c(1:length(y3))

sv_bpf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 3))[,1]
sv_lw3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 4))[,1]
sv_bapf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 6))[,1]
dfsv13<-data.frame(timeframe,y3,sv_bpf3)
dfsv14<-data.frame(timeframe,y3,sv_gpfopt3)
dfsv15<-data.frame(timeframe,y3,sv_apf3)
dfsv16<-data.frame(timeframe,y3,sv_lw3)
dfsv17<-data.frame(timeframe,y3,sv_bapf3)
dfsv18<-data.frame(timeframe,y3,sv_sis3)

Filtplot13<-function(dataframe,title){
  mean=sv_bpf3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot14<-function(dataframe,title){
  mean=sv_gpfopt3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot15<-function(dataframe,title){
  mean=sv_apf3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot16<-function(dataframe,title){
  mean=sv_lw3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot17<-function(dataframe,title){
  mean=sv_bapf3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot18<-function(dataframe,title){
  mean=sv_sis3
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Observations" = "gray",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Observations" = 1,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}

plotobs13<-Filtplot13(dfsv13,"Bootstrap Particle Filter DJI")
plotobs14<-Filtplot14(dfsv14,"Opt Ker Guided Particle Filter DJI")
plotobs15<-Filtplot15(dfsv15,"Auxiliary Particle Filter DJI")
plotobs16<-Filtplot16(dfsv16,"Liu and West Filter DJI")
plotobs17<-Filtplot17(dfsv17,"Always Resampling DJI")
plotobs18<-Filtplot18(dfsv18,"No Resampling DJI")

ggarrange(plotobs13,plotobs14)
ggarrange(plotobs15,plotobs16)
ggarrange(plotobs17,plotobs18)

First, a general point is that the sequence of percentage log returns are different across the three indices. The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents. In particular, as previously noticed, there are extensive similarities between the DOW JONES Index and the S&P500, which can be visualized when looking at the percentage of log returns in the corresponding plots, since both are describing the US Economy. However, the STOXX50E differs quite extensively, especially in the different sensitivity to the Covid-19 Crisis and its consequences,  representing the European Economy. Aside from this note, the considerations expressed below on the sequences "filtered volatilities" for the various techniques hold almost uniformly for all the three indices. Therefore, in the next paragraph we will be referring to the set of plots pertaining to one specific index (which could be any of the three).\

@@

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Riferimento alla sezione di convergenza?.} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500. Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}

We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark. In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles. Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below. Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling algorithm. By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never). In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms. Finally, note that the Auxiliary Particle Filter employed here does not coincide with that of the section with synthetic data, which employed a different kernel implied by optimality results.}

@@

As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse. With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.\
As it turns out, such suggestion is confirmed by running the algorithms with N=1000. Passing from 100 to 1000 particles more than halves the RMSEs across the 5 competing filters, with similarly pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a shrinking hedge over the Auxiliary PF, which shows most impressive gains when passing from 100 to 1000 particles. Further sizable gains for every filter are observed when increasing the number of particles to 5000, a level at which both the RMSE and the MAE agree on a slight hedge of the Guided and the Bootstrap filters over the Auxiliary and the Basic PFs.\
Nevertheless, we would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others, a figure which visual inspection of the graphs below can confirm. Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (note that the Basic PF is preferred to the APF only twice, by the RMSE), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles. Moreover, the RMSE and the MAE show infrequent disagreement, delivering a rather clear picture of the comparison.

```{r echo=FALSE, fig.align='center', fig.cap="Approximations at N=10000, Bootstrap benchmark", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}

# Benchmark comparison Plots

realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
y<-realdatasetSPX[,2]
btbenchmark<-(read.xlsx("svbpf50000p.xlsx", sheet = 1))[,1]
timeframe<-c(1:length(y))

sv_bpf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,1]
sv_bapf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,1]
sv_bpf1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,2]
sv_gpfopt1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,2]
sv_apf1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,2]
sv_bapf1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,2]
sv_sis1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,2]

dfsv1<-data.frame(timeframe,y,btbenchmark,sv_bpf1,sv_bpf1sd)
dfsv2<-data.frame(timeframe,y,btbenchmark,sv_gpfopt1,sv_gpfopt1sd)
dfsv3<-data.frame(timeframe,y,btbenchmark,sv_apf1,sv_apf1sd)
dfsv5<-data.frame(timeframe,y,btbenchmark,sv_bapf1,sv_bapf1sd)
dfsv6<-data.frame(timeframe,y,btbenchmark,sv_sis1,sv_sis1sd)

Filtplot1<-function(dataframe,title){
  mean=sv_bpf1
  sd=sv_bpf1sd
  dataframe<-data.frame(dataframe,mean)
